Re-analysing the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): A textbook illustration of the absence of evidence fallacy

### 2020-10-08 manuscript.Rmd
updating the dance plot

Once again, anticipating the legitimate critique that the absence of a significant difference is not necessarily "significant" evidence for the absence of an effect, @moffatt_inner_2020 reported the following Bayes factor (BF) analysis:

@@

```{r bf-dance, fig.cap = "Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017)."}
# inspired from http://daniellakens.blogspot.com/2016/07/dance-of-bayes-factors.html
# population effect size
cohensd <- 0.36

# sample size
n <- 26

@@

# plotting the results
results %>%
  pivot_longer(cols = BF10:BF10_double, names_to = "type") %>%
  mutate(
    type = factor(
      type,
      levels = c("BF10_null", "BF10", "BF10_double"),
      labels = c("Cohen's d = 0", "Cohen's d = 0.36", "Cohen's d = 0.72")
      )
    ) %>%
  ggplot(aes(x = log(value), fill = type) ) +
  geom_density(alpha = 0.5, show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_bw(base_size = 12) +
  facet_wrap(~ type, scales = "free_x") +
  labs(x = "Natural logarithm of the BF10", y = "Density") +
  scale_colour_brewer(palette = "Dark2", direction = 1) +
  scale_fill_brewer(palette = "Dark2", direction = 1)

@@

First, we simulated 10.000 datasets (for $N = 26$) under the assumption of either no effect (i.e., the null hypothesis of $d = 0$), an effect size of $d = 0.36$ [i.e., the supposed target effect size in @moffatt_inner_2020], or an effect size of $d = 0.72$ [i.e., the effect size reported in @nalborczyk_orofacial_2017]. As shown in Figure \@ref(fig:bf-dance), the distribution of BFs computed under each hypothesis reveals important inter-simulation variability. For instance, under the null hypothesis, `r round(prop_d0 * 100, 2)`% of the computed BFs are above 0 and hence support the alternative hypothesis (although the "true" effect size is $d = 0$). When the "true" effect size is of $d = 0.36$, `r round(100 - prop_d0.36 * 100, 2)`% of the BFs are below 0 and hence support the null hypothesis (although the true effect size is actually non-null). When the "true" effect size is of $d = 0.72$, `r round(100 - prop_d0.72 * 100, 2)`% of the BFs are still below 0. In other words, for small sample and effect sizes, BFs have high error rates.^[To assess the extent to which the $\text{BF}_{10}$ computed for the OOI by @moffatt_inner_2020 is "surprising" given or "compatible" with the hypothesis of an effect size of $d = 0.36$, we can compute the probability of obtaining this finding or a more extreme finding given the hypothesis, which is approximately equal to `r round(1 - mean(results$BF10 > 0.278), 2)`.]

Second, we conducted another simulation with the aim of assessing the relation between the sample size and the value of the BF. In the previous section, we fitted a multivariate Bayesian regression model with varying-intercepts by participant and weakly informative priors on the EMG data collected by @moffatt_inner_2020. Using this model, we i) generated new datasets from the posterior predictive distribution of this model and ii) we computed the BF in favour of the alternative hypothesis ($\text{BF}_{10}$) using the `BayesFactor` package [@R-BayesFactor]. We used a "medium" prior (i.e., $r = \sqrt{2}/2$) on the scale of the Cauchy prior for the alternative hypothesis. We repeated this procedure for varying sample sizes from 20 to 200 participants (by increments of 10 participants) with $1000$ simulations (i.e., 1000 simulated datasets) for each sample size.

@@

> "Unless otherwise reported, the inclusion of order in which the conditions were completed as a between-subjects variable as part of a mixed-design ANOVA produced no significant main effects or interactions involving order."

@@

To conclude, we wish to bring some nuance to the conclusion of @moffatt_inner_2020, who stated that "In conclusion, induced rumination appeared to involve similar levels of inner speech-related muscle activity to a period of distraction". In consideration of the limitations discussed in the present article, this conclusion seems hasty. Indeed, we provided theoretical (epistemological) and empirical (via simulation) reasons to doubt the strength of the evidence for the null hypothesis in this study. Moreover, supplementary analyses showed that the order of the conditions participants went through may have influenced the effects of the rumination induction on the EMG correlates. Finally, important under-explored inter-individual variability suggests that important determinants of these correlates were not taken into account. We urge the authors to nuance their conclusions, to analyse further their data, and to plan adequately-powered studies in order to settle these issues.

@@

We thank Antonio Schettino for suggesting to include the "Dance of the Bayes factors" simulation and for provding helpful comments on a previous version of this manuscript.

>>>>

Once again, anticipating the legitimate critique that the absence of a significant difference is not necessarily "significant" evidence for the absence of an effect, @moffatt_inner_2020 reported the following Bayes factor (BF) analysis (p.12):

@@

```{r bf-dance, fig.cap = "Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017). The red vertical dashed line indicates the value of the BF computed for the OOI by Moffatt et al. (2020), on the log scale."}
# inspired from http://daniellakens.blogspot.com/2016/07/dance-of-bayes-factors.html
# population effect size
cohensd <- 0.36

# sample size in Mofatt et al. (2020)
n <- 26

@@

# plotting the results
results %>%
  pivot_longer(cols = BF10:BF10_double, names_to = "type") %>%
  mutate(
    type = factor(
      type,
      levels = c("BF10_null", "BF10", "BF10_double"),
      labels = c("Cohen's d = 0", "Cohen's d = 0.36", "Cohen's d = 0.72")
      )
    ) %>%
  ggplot(aes(x = log(value), fill = type) ) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_vline(xintercept = log(0.278), linetype = 2, colour = "orangered") +
  geom_histogram(colour = "white", show.legend = FALSE, bins = 50) +
  theme_bw(base_size = 12) +
  facet_wrap(~ type, scales = "free_x") +
  labs(x = "Natural logarithm of the BF10", y = "Number of simulations") +
  scale_colour_brewer(palette = "Dark2", direction = 1) +
  scale_fill_brewer(palette = "Dark2", direction = 1)

@@

First, we simulated 10.000 datasets (for $N = 26$) under the assumption of either no effect (i.e., the null hypothesis of $d = 0$), an effect size of $d = 0.36$ [i.e., the supposed target effect size in @moffatt_inner_2020], or an effect size of $d = 0.72$ [i.e., the effect size reported in @nalborczyk_orofacial_2017]. As shown in Figure \@ref(fig:bf-dance), the distribution of BFs computed under each hypothesis reveals important inter-simulation variability. For instance, under the null hypothesis, `r round(prop_d0 * 100, 2)`% of the computed BFs are above 0 and hence support the alternative hypothesis (although the "true" effect size is $d = 0$). When the "true" effect size is of $d = 0.36$, `r round(100 - prop_d0.36 * 100, 2)`% of the BFs are below 0 and hence support the null hypothesis (although the true effect size is actually non-null). When the "true" effect size is of $d = 0.72$, `r round(100 - prop_d0.72 * 100, 2)`% of the BFs are still below 0. In other words, for small sample and effect sizes, BFs have high error rates.^[To assess the extent to which the BF computed for the OOI by @moffatt_inner_2020 (i.e., $\text{BF}_{10} = 0.278$) is "surprising" given or "compatible" with the hypothesis of an effect size of $d = 0.36$, we can compute the probability of obtaining this finding or a more extreme finding given the hypothesis, which is approximately equal to `r round(1 - mean(results$BF10 > 0.278), 2)`.]

Second, we conducted another simulation with the aim of assessing the relation between the sample size and the value of the BF. In the previous section, we fitted a multivariate Bayesian regression model with varying-intercepts by participant and weakly informative priors on the EMG data collected by @moffatt_inner_2020. Using this model, we i) generated new datasets from the posterior predictive distribution of this model and ii) we computed the BF in favour of the alternative hypothesis ($\text{BF}_{10}$) using the `BayesFactor` package [@R-BayesFactor]. We used a "medium" prior (i.e., $r = \sqrt{2}/2$) on the scale parameter of the Cauchy prior for the alternative hypothesis. We repeated this procedure for varying sample sizes from 20 to 200 participants (by increments of 10 participants) with $1000$ simulations (i.e., 1000 simulated datasets) for each sample size.

@@

> "Unless otherwise reported, the inclusion of order in which the conditions were completed as a between-subjects variable as part of a mixed-design ANOVA produced no significant main effects or interactions involving order." (p.7)

@@

To conclude, we wish to bring some nuance to the conclusion of @moffatt_inner_2020, who stated that "In conclusion, induced rumination appeared to involve similar levels of inner speech-related muscle activity to a period of distraction" (p.14). In consideration of the limitations discussed in the present article, this conclusion seems hasty. Indeed, we provided theoretical (epistemological) and empirical (via simulation) reasons to doubt the strength of the evidence for the null hypothesis in this study. Moreover, supplementary analyses showed that the order of the conditions participants went through may have influenced the effects of the rumination induction on the EMG correlates. Finally, important under-explored inter-individual variability suggests that important determinants of these correlates were not taken into account. We urge the authors to nuance their conclusions, to analyse further their data, and to plan adequately-powered studies in order to settle these issues.

@@

We thank Antonio Schettino for suggesting to include the "Dance of the Bayes factors" simulation and for providing helpful comments on a previous version of this manuscript.

### 2020-10-08 manuscript.tex
updating the dance plot

Wordcount (excluding abstract, references, tables, and figures): 4549

@@

Once again, anticipating the legitimate critique that the absence of a significant difference is not necessarily ``significant'' evidence for the absence of an effect, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the following Bayes factor (BF) analysis:

@@

\caption{Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017).}\label{fig:bf-dance}
\end{figure}

First, we simulated 10.000 datasets (for \(N = 26\)) under the assumption of either no effect (i.e., the null hypothesis of \(d = 0\)), an effect size of \(d = 0.36\) (i.e., the supposed target effect size in Moffatt et al., \protect\hyperlink{ref-moffatt_inner_2020}{2020}), or an effect size of \(d = 0.72\) (i.e., the effect size reported in Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}). As shown in Figure \ref{fig:bf-dance}, the distribution of BFs computed under each hypothesis reveals important inter-simulation variability. For instance, under the null hypothesis, 6.96\% of the computed BFs are above 0 and hence support the alternative hypothesis (although the ``true'' effect size is \(d = 0\)). When the ``true'' effect size is of \(d = 0.36\), 71.83\% of the BFs are below 0 and hence support the null hypothesis (although the true effect size is actually non-null). When the ``true'' effect size is of \(d = 0.72\), 25.58\% of the BFs are still below 0. In other words, for small sample and effect sizes, BFs have high error rates.\footnote{To assess the extent to which the \(\text{BF}_{10}\) computed for the OOI by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is ``surprising'' given or ``compatible'' with the hypothesis of an effect size of \(d = 0.36\), we can compute the probability of obtaining this finding or a more extreme finding given the hypothesis, which is approximately equal to 0.29.}

Second, we conducted another simulation with the aim of assessing the relation between the sample size and the value of the BF. In the previous section, we fitted a multivariate Bayesian regression model with varying-intercepts by participant and weakly informative priors on the EMG data collected by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}). Using this model, we i) generated new datasets from the posterior predictive distribution of this model and ii) we computed the BF in favour of the alternative hypothesis (\(\text{BF}_{10}\)) using the \texttt{BayesFactor} package (Morey \& Rouder, \protect\hyperlink{ref-R-BayesFactor}{2018}). We used a ``medium'' prior (i.e., \(r = \sqrt{2}/2\)) on the scale of the Cauchy prior for the alternative hypothesis. We repeated this procedure for varying sample sizes from 20 to 200 participants (by increments of 10 participants) with \(1000\) simulations (i.e., 1000 simulated datasets) for each sample size.

@@

\begin{quote}
``Unless otherwise reported, the inclusion of order in which the conditions were completed as a between-subjects variable as part of a mixed-design ANOVA produced no significant main effects or interactions involving order.''
\end{quote}

@@

To conclude, we wish to bring some nuance to the conclusion of Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), who stated that ``In conclusion, induced rumination appeared to involve similar levels of inner speech-related muscle activity to a period of distraction''. In consideration of the limitations discussed in the present article, this conclusion seems hasty. Indeed, we provided theoretical (epistemological) and empirical (via simulation) reasons to doubt the strength of the evidence for the null hypothesis in this study. Moreover, supplementary analyses showed that the order of the conditions participants went through may have influenced the effects of the rumination induction on the EMG correlates. Finally, important under-explored inter-individual variability suggests that important determinants of these correlates were not taken into account. We urge the authors to nuance their conclusions, to analyse further their data, and to plan adequately-powered studies in order to settle these issues.

@@

We thank Antonio Schettino for suggesting to include the ``Dance of the Bayes factors'' simulation and for provding helpful comments on a previous version of this manuscript.

>>>>

Wordcount (excluding abstract, references, tables, and figures): 4561

@@

Once again, anticipating the legitimate critique that the absence of a significant difference is not necessarily ``significant'' evidence for the absence of an effect, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the following Bayes factor (BF) analysis (p.12):

@@

\caption{Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017). The red vertical dashed line indicates the value of the BF computed for the OOI by Moffatt et al. (2020), on the log scale.}\label{fig:bf-dance}
\end{figure}

First, we simulated 10.000 datasets (for \(N = 26\)) under the assumption of either no effect (i.e., the null hypothesis of \(d = 0\)), an effect size of \(d = 0.36\) (i.e., the supposed target effect size in Moffatt et al., \protect\hyperlink{ref-moffatt_inner_2020}{2020}), or an effect size of \(d = 0.72\) (i.e., the effect size reported in Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}). As shown in Figure \ref{fig:bf-dance}, the distribution of BFs computed under each hypothesis reveals important inter-simulation variability. For instance, under the null hypothesis, 6.96\% of the computed BFs are above 0 and hence support the alternative hypothesis (although the ``true'' effect size is \(d = 0\)). When the ``true'' effect size is of \(d = 0.36\), 71.83\% of the BFs are below 0 and hence support the null hypothesis (although the true effect size is actually non-null). When the ``true'' effect size is of \(d = 0.72\), 25.58\% of the BFs are still below 0. In other words, for small sample and effect sizes, BFs have high error rates.\footnote{To assess the extent to which the BF computed for the OOI by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) (i.e., \(\text{BF}_{10} = 0.278\)) is ``surprising'' given or ``compatible'' with the hypothesis of an effect size of \(d = 0.36\), we can compute the probability of obtaining this finding or a more extreme finding given the hypothesis, which is approximately equal to 0.29.}

Second, we conducted another simulation with the aim of assessing the relation between the sample size and the value of the BF. In the previous section, we fitted a multivariate Bayesian regression model with varying-intercepts by participant and weakly informative priors on the EMG data collected by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}). Using this model, we i) generated new datasets from the posterior predictive distribution of this model and ii) we computed the BF in favour of the alternative hypothesis (\(\text{BF}_{10}\)) using the \texttt{BayesFactor} package (Morey \& Rouder, \protect\hyperlink{ref-R-BayesFactor}{2018}). We used a ``medium'' prior (i.e., \(r = \sqrt{2}/2\)) on the scale parameter of the Cauchy prior for the alternative hypothesis. We repeated this procedure for varying sample sizes from 20 to 200 participants (by increments of 10 participants) with \(1000\) simulations (i.e., 1000 simulated datasets) for each sample size.

@@

\begin{quote}
``Unless otherwise reported, the inclusion of order in which the conditions were completed as a between-subjects variable as part of a mixed-design ANOVA produced no significant main effects or interactions involving order.'' (p.7)
\end{quote}

@@

To conclude, we wish to bring some nuance to the conclusion of Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), who stated that ``In conclusion, induced rumination appeared to involve similar levels of inner speech-related muscle activity to a period of distraction'' (p.14). In consideration of the limitations discussed in the present article, this conclusion seems hasty. Indeed, we provided theoretical (epistemological) and empirical (via simulation) reasons to doubt the strength of the evidence for the null hypothesis in this study. Moreover, supplementary analyses showed that the order of the conditions participants went through may have influenced the effects of the rumination induction on the EMG correlates. Finally, important under-explored inter-individual variability suggests that important determinants of these correlates were not taken into account. We urge the authors to nuance their conclusions, to analyse further their data, and to plan adequately-powered studies in order to settle these issues.

@@

We thank Antonio Schettino for suggesting to include the ``Dance of the Bayes factors'' simulation and for providing helpful comments on a previous version of this manuscript.
